Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update review discussion

# Discussion
## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
Severe drought events negatively affected to primary and secondary growth of *Q. pyrenaica* forests in the rear edge of their distribution. Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests. These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. 
Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge.

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.g.* first months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites). This could explain the higher values of recovery found for 1995 (Figure 6). The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*).

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to contextualize remote sensing data by reflecying climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. Combined anlaysis, such the approach used in our study, may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [Babst2018].

## Microenvironment shapes differential sensitivity to climate and drought of rear-edge oaks

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). The former, which have drier conditions (Table 1), were more impacted by the drought events than the latter. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in oaks [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.  [@Bhuyan2017].

@@

In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).

## Land-use legacies shape distribution and sensitivity to climate change at the read-edge

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents (Table S4). As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].

Our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4), driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015] (Table S4). On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@JimenezOlivencia2015]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops. Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure. For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history (Table S4). On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation [@Calatrava2019]. Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013]. In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment. Acorn resources have been tradiotionally exploited in this location up to 1950s (Tabla S4).

Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area (1925 to 1957) (Table S4). During this period there were an increase on the use of timbers for the tunnels of the mines and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area [@Titos1990]. This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017]. The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995 (*personal communications*) as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

For southern sites, the only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Navarro2013; @DoblasMiranda2017; @Penuelas2017] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species. Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].

# Conclussions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019]. The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.

>>>>

# Discussion
## Relict oaks show high resilience to recent drought events and long-term climatic variability

Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. However, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. In addition, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species [@SanchezSalguero2012; @Camarero2015]. But, interestingly, the highest values of resilience for tree-growth were obtained for these two drought events. In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6). This result seems contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites. These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].

## Small-scale environmental variability shapes the sensitivity to climate of trees within the rear-edge

Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event. In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site, can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014]. Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017].

It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017]. It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Remarkably, the trees at CA-High, which are located around 1900 m. *a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4). This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species. The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017]. Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].

Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).

@@

## Land-use legacies shape forests and sensitivity to climate change at the rear-edge

The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with a sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, firewood extraction and charcoal exploitations (these were strongly suggested in the disturbance chronology in SJ), pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015]. Therefore, our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation. Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between Northern (coppice) and Southern populations (high forest, open woodland) [@JimenezOlivencia2015]. It has been described a sequentially distribution of the land uses along the elevational gradient in the northern site (*e.g.* San Juan): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015]. Other activities (like mining) should have determined the forest structure at that northern location, sincr SJ is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history. On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands along the elevational gradient where oaks occurr [@Calatrava2019], being firewood, charcoal and acorns continuously exploited through history [@Valbuena2013]. Charcoal extraction is weel known at least since the 15th century up to middle of the 20th century [@JimenezSerrano2004]. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment and use of gas and fosil fuels, similar to other Fagaceae stands across Europe and the Mediterranean [@Valbuena2013].

Some of the previous historical records of land-use were expressed in our growth chronologies. The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4). In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in the northern site [@GarciaAlix2017]. The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). Therefore, two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

For southern sites, the only positive release event showed at the earliest yers for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].

... revisar 
Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].  
...

# Conclusions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019]. The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
